---
title: "Latent time joint mixed effect model simulation & fit"
author: "Dan Li and  Michael C. Donohue"
output: 
  rmarkdown::html_document:
    toc: true
    toc_float: true
vignette: >
  %\VignetteIndexEntry{Latent time joint mixed effect model simulation and fit}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r knitr_settings, results = 'hide', echo = FALSE, message = FALSE, cache = FALSE, warning = FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE, tidy = FALSE, 
  fig_width = 6.5, fig_height = 4, comment = NA)
```

```{r libraries}
library(dplyr)
library(tidyr)
library(tibble)
library(ggplot2)
library(ltjmm)
library(bayesplot)
library(posterior)
```

```{r settings, echo=FALSE}
theme_set(theme_bw(base_size = 12))
cbbPalette <- c("#0072B2", "#D55E00", "#E69F00", 
  "#F0E442", "#56B4E9", "#CC79A7", "#009E73", "#999999", "#000000")
scale_colour_discrete <- function(...) scale_colour_manual(..., values=cbbPalette)
scale_fill_discrete <- function(...) scale_fill_manual(..., values=cbbPalette)
```

This vignettes demonstrates simulating and fitting latent time joint mixed effect models as described in Li, et al. 2017\footnote{Li, D., Iddi, S., Thompson, W. K., Donohue, M. C., for the Alzheimer's Disease Neuroimaging Initiative. (2017). Bayesian latent time joint mixed effect models for multicohort longitudinal data. \emph{Statistical methods in medical research}, \url{https://doi.org/10.1177/0962280217737566}.}.

The simulated data includes latent times. Random intercepts and slopes are simulated 

# Simulate one example dataset
```{r simulationParameters}
rng_seed <- 20161001
set.seed(rng_seed)
n <- 400   # subjects
p <- 4     # outcomes
t <- 4     # time points

dd <- tibble(
    id = 1:n,
    age.0 = rnorm(n, 75, 5)) %>%
  cross_join(tibble(visit = 1:t)) %>%
  mutate(year = (visit-1)*3) %>%
  cross_join(tibble(outcome = 1:p)) %>%
  arrange(id, outcome, visit) %>%
  mutate(
    Outcome = factor(outcome, levels = 1:p, labels = paste('Outcome', 1:p)),
    Y = NA)

setup_ltjmm <- ltjmm(Y ~ year | 1 | id | outcome, data = dd)

## variance parameters 
sigma_y <- c(0.4, 0.4, 0.6, 0.5)
# one less degree of freedom for intercepts due to identifiability constraint:
sigma_alpha0 <- c(0.5, 1, 0.8) 
sigma_alpha1 <- c(0.1, 0.1, 0.1, 0.1)
sigma_alpha <- c(sigma_alpha0, sigma_alpha1)
sigma_delta <- 4
N_X <- 1
beta <- matrix(c(1, 0.5, 2, 0.8), p, N_X)
gamma <- c(0.2, 0.3, 0.25, 0.5)

simulated_ltjmm <- simulate(setup_ltjmm,
  beta = beta,
  gamma = gamma,
  sigma_diag = sigma_alpha,
  sigma_delta = sigma_delta,
  sigma_y = sigma_y,
  seed = 201610014)

dd$Y <- simulated_ltjmm$y

ECDF.list <- lapply(unique(dd$outcome), function(oc){
  with(subset(dd, outcome == oc), ecdf(Y))
})

IECDF.list <- lapply(unique(dd$outcome), function(oc){
  inv.ecdf.data <- with(subset(dd, outcome == oc),
    as.numeric(quantile(Y, probs=seq(0,1,by=0.01))))
  approxfun(seq(0,1,by=0.01), inv.ecdf.data)
})

dd <- dd %>% mutate(
  Q = case_when(
    outcome == 1 ~ ECDF.list[[1]](Y),
    outcome == 2 ~ ECDF.list[[2]](Y),
    outcome == 3 ~ ECDF.list[[3]](Y),
    outcome == 4 ~ ECDF.list[[4]](Y),
  ),
  Z = qnorm(Q)) %>%
  left_join(tibble(
    id = 1:n,
    delta_true = simulated_ltjmm$delta), by = 'id')
```

```{r, fig.cap="Simulated outcomes relative to unkown time shift, delta, on the raw scale."}
ggplot(dd, aes(x=year+delta_true, y=Y)) +
  geom_line(aes(group=id), alpha =0.2) +
  facet_wrap(~Outcome)
```

```{r, fig.cap="Simulated outcomes relative to unkown time shift, delta, on the quantile scale."}
ggplot(dd, aes(x=year+delta_true, y=Q)) +
  geom_line(aes(group=id), alpha =0.2) +
  facet_wrap(~Outcome)
```

```{r, fig.cap="Simulated outcomes relative to unkown time shift, delta, on the standard normal scale."}
ggplot(dd, aes(x=year+delta_true, y=Z)) +
  geom_line(aes(group=id), alpha =0.2) +
  facet_wrap(~Outcome)
```

# Fit LTJMM with independent random effects with Stan
```{r stanFit, eval=FALSE}
fit <- ltjmm_stan(Y ~ year |
    1  | # fixed effects direct on outcome
    id | outcome,
  data = dd,
  seed = rng_seed,
  chains = 2, parallel_chains = 2,
  iter_warmup = 500, iter_sampling = 500, thin = 2)
# fit$save_object(file = 'sim_results.RDS')
fit.sum <- fit$summary(c("beta", "gamma", "sigma_delta", "delta", "alpha0", "alpha1")) %>%
  as.data.frame()
row.names(fit.sum) <- fit.sum %>% pull(variable)
draws.beta <- fit$draws("beta")
draws.gamma <- fit$draws("gamma")
draws.sigma_delta <- fit$draws("sigma_delta")
save(fit.sum, draws.beta, draws.gamma, draws.sigma_delta, 
  file = 'sim_results.rdata')
```

```{r loadResults}
load(system.file("testdata", "sim_results.rdata", package="ltjmm"))
```

## Diagnostic plots
```{r traceplotBeta, fig_width = 6.5, fig_height = 4}
mcmc_trace(draws.beta)
```

```{r traceplotGamma}
mcmc_trace(draws.gamma)
```

```{r traceplotSigmaDelta}
mcmc_trace(draws.sigma_delta)
```

## True versus posterior mean of latent time shifts
```{r biasDeltaPlot}
delta <- simulated_ltjmm$delta
delta.posteriormean <- fit.sum %>% filter(grepl('delta', variable)) %>% 
  filter(variable != 'sigma_delta') %>%
  pull(mean)
par(mgp = c(2.2, 0.45, 0), tcl = -0.4, mar = c(3.3, 3.6, 1.1, 1.1))
plot(delta, delta.posteriormean,
  xlim = range(delta), ylim = range(delta), 
  xlab = expression(paste("True value of time shift ", delta[i])),
  ylab = expression(paste("Posterior mean of time shift ", delta[i])))
abline(0, 1, lwd=2, col='red', lty = 2)
```

## True versus posterior mean of random intercepts and slopes for each outcome
```{r biasAlphaPlot}
alpha0true <- as.data.frame(simulated_ltjmm$alpha0) %>% mutate(id = 1:n, parameter='alpha0') %>% 
  gather(outcome, truth, V1:V4)
alpha1true <- as.data.frame(simulated_ltjmm$alpha1) %>% mutate(id = 1:n, parameter='alpha1') %>% 
  gather(outcome, truth, V1:V4)
alphapm <- data.frame(
  parameter.id.outcome = grep('alpha', row.names(fit.sum), value=TRUE), 
  posterior.mean = fit.sum[grepl('alpha', row.names(fit.sum)), 'mean']) %>% 
  separate(parameter.id.outcome, c('parameter', 'id', 'outcome', 'other')) %>% 
  mutate(id = as.numeric(id))

pd <- full_join(alpha0true, alpha1true, by=c('id', 'outcome', 'parameter', 'truth')) %>%
  mutate(outcome = gsub('V', '', outcome)) %>% 
  full_join(alphapm) 
ggplot(pd, aes(x=truth, y=posterior.mean)) + 
  geom_point(aes(shape=parameter, color=parameter), alpha=0.25) +
  geom_abline(intercept=0, slope=1) + facet_wrap(~outcome, scales='free')
```

## Model prediction plots

```{r}
nd0 <- expand_grid(
  `(Intercept)` = 1,
  `year+delta` = seq(-10, 20, by=1),
  outcome = 1:p) %>%
  cross_join(as_draws_df(draws.beta)) %>%
  left_join(as_draws_df(draws.gamma), by = c('.iteration', '.chain', '.draw')) %>%
  mutate(
    Y = case_when(
      outcome == 1 ~ 
        `year+delta`*`gamma[1]` + `beta[1,1]`,
      outcome == 2 ~ 
        `year+delta`*`gamma[2]` + `beta[2,1]`,
      outcome == 3 ~ 
        `year+delta`*`gamma[3]` + `beta[3,1]`,
      outcome == 4 ~ 
        `year+delta`*`gamma[4]` + `beta[4,1]`),
    Q = case_when(
      outcome == 1 ~ ECDF.list[[1]](Y),
      outcome == 2 ~ ECDF.list[[2]](Y),
      outcome == 3 ~ ECDF.list[[3]](Y),
      outcome == 4 ~ ECDF.list[[4]](Y)))

nd <- nd0 %>%
  group_by(outcome, `year+delta`) %>%
  summarise(
    median_Y = median(Y),
    lower.CL_Y = quantile(Y, probs = 0.025)[1],
    upper.CL_Y = quantile(Y, probs = 1-0.025)[1],
    median_Q = median(Q),
    lower.CL_Q = quantile(Q, probs = 0.025)[1],
    upper.CL_Q = quantile(Q, probs = 1-0.025)[1]) %>%
  pivot_longer(median_Y:upper.CL_Q) %>%
  separate(name, into = c('stat', 'scale'), sep = '_') %>%
  pivot_wider(names_from = stat, values_from = value) %>%
  mutate(
    Outcome = factor(outcome, levels = c(1, 2, 3, 4), 
      labels = paste('Outcome', 1:p)))

dd <- dd %>%
  left_join(
    tibble(id = 1:n, delta_hat = delta.posteriormean), by = 'id')
```

```{r, fig.cap='Model predictions and observations on the original scale relative to estimated time shifts.'}
ggplot(filter(nd, scale == 'Y'), aes(x=`year+delta`, y=median)) +
  geom_line(aes(color = Outcome)) +
  facet_wrap(~Outcome) +
  geom_ribbon(aes(ymin=lower.CL, ymax=upper.CL, fill=Outcome), alpha=0.2) +
  geom_point(data = dd, aes(x = year + delta_hat, y = Y), alpha = 0.01) +
  ylab('Y')
```

```{r, fig.cap='Model predictions and observations on the quantile scale relative to estimated time shifts.'}
ggplot(filter(nd, scale == 'Q'), aes(x=`year+delta`, y=median)) +
  geom_line(aes(color = Outcome)) +
  facet_wrap(~Outcome) +
  geom_ribbon(aes(ymin=lower.CL, ymax=upper.CL, fill=Outcome), alpha=0.2) +
  geom_point(data = dd, aes(x = year + delta_hat, y = Q), alpha = 0.01) +
  ylab('Q')
```

```{r, fig.cap='Model predictions and 95% credible intervals on the quantile scale relative to estimated time shifts.'}
ggplot(filter(nd, scale == 'Q'), aes(x=`year+delta`, y=median)) +
  geom_line(aes(color = Outcome)) +
  geom_ribbon(aes(ymin=lower.CL, ymax=upper.CL, fill=Outcome), alpha=0.2) +
  ylab('Q')
```

# Fit LTJMM with multivariate random effects with Stan
```{r stanFitMulti, eval=FALSE}
fit <- ltjmm_stan(Y ~ year |
    1  | # fixed effects direct on outcome
    id | outcome,
  random_effects = "multivariate",
  data = dd,
  seed = rng_seed,
  chains = 2, parallel_chains = 2,
  iter_warmup = 500, iter_sampling = 500, thin = 2)
# fit$save_object(file = 'sim_lt_multi_results.RDS')
fit.sum <- fit$summary(c("beta", "gamma", "sigma_delta", "delta", "alpha0", "alpha1")) %>%
  as.data.frame()
row.names(fit.sum) <- fit.sum %>% pull(variable)
draws.beta <- fit$draws("beta")
draws.gamma <- fit$draws("gamma")
draws.sigma_delta <- fit$draws("sigma_delta")
save(fit.sum, draws.beta, draws.gamma, draws.sigma_delta, 
  file = 'sim_lt_multi_results.rdata')
```

```{r loadResultsMulti}
load(system.file("testdata", "sim_lt_multi_results.rdata", package="ltjmm"))
```

## Diagnostic plots
```{r traceplotBetaMulti}
mcmc_trace(draws.beta)
```

```{r traceplotGammaMulti}
mcmc_trace(draws.gamma)
```

## True versus posterior mean of latent time shifts
```{r biasDeltaPlotMulti}
delta <- simulated_ltjmm$delta
delta.posteriormean <- fit.sum %>% filter(grepl('delta', variable)) %>% 
  filter(variable != 'sigma_delta') %>%
  pull(mean)
par(mgp = c(2.2, 0.45, 0), tcl = -0.4, mar = c(3.3, 3.6, 1.1, 1.1))
plot(delta, delta.posteriormean,
  xlim = range(delta), ylim = range(delta), 
  xlab = expression(paste("True value of time shift ", delta[i])),
  ylab = expression(paste("Posterior mean of time shift ", delta[i])))
abline(0, 1, lwd=2, col='red', lty = 2)
```

## True versus posterior mean of random intercepts and slopes for each outcome
```{r biasAlphaPlotMulti}
alpha0true <- as.data.frame(simulated_ltjmm$alpha0) %>% mutate(id = 1:n, parameter='alpha0') %>% 
  gather(outcome, truth, V1:V4)
alpha1true <- as.data.frame(simulated_ltjmm$alpha1) %>% mutate(id = 1:n, parameter='alpha1') %>% 
  gather(outcome, truth, V1:V4)
alphapm <- data.frame(
  parameter.id.outcome = grep('alpha', row.names(fit.sum), value=TRUE), 
  posterior.mean = fit.sum[grepl('alpha', row.names(fit.sum)), 'mean']) %>% 
  separate(parameter.id.outcome, c('parameter', 'id', 'outcome', 'other')) %>% 
  mutate(id = as.numeric(id))

pd <- full_join(alpha0true, alpha1true, by=c('id', 'outcome', 'parameter', 'truth')) %>%
  mutate(outcome = gsub('V', '', outcome)) %>% 
  full_join(alphapm) 
ggplot(pd, aes(x=truth, y=posterior.mean)) + 
  geom_point(aes(shape=parameter, color=parameter), alpha=0.25) +
  geom_abline(intercept=0, slope=1) + facet_wrap(~outcome, scales='free')
```

# Fit joint mixed effect model (JMM) with multivariate random effects with Stan
```{r stanFitJMM, eval=FALSE}
fit <- ltjmm_stan(Y ~ year |
    1  | # fixed effects direct on outcome
    id | outcome,
  lt = FALSE,
  random_effects = "multivariate",
  data = dd,
  seed = rng_seed,
  chains = 2, parallel_chains = 2,
  iter_warmup = 250, iter_sampling = 250, thin = 1)
# fit$save_object(file = 'sim_jmm_results.RDS')
fit.sum <- fit$summary(c("beta", "gamma", "alpha0", "alpha1")) %>%
  as.data.frame()
row.names(fit.sum) <- fit.sum %>% pull(variable)
draws.beta <- fit$draws("beta")
draws.gamma <- fit$draws("gamma")
save(fit.sum, draws.beta, draws.gamma, 
  file = 'sim_jmm_results.rdata')
```

```{r loadResultsJMM}
load(system.file("testdata", "sim_jmm_results.rdata", package="ltjmm"))
```

## Diagnostic plots
```{r traceplotBetaJMM, fig.cap='Note. Model is mispecified, which might explain poor mixing.'}
mcmc_trace(draws.beta)
```

## True versus posterior mean of random intercepts and slopes for each outcome
```{r biasAlphaPlotJMM, fig.cap='Note. Model is mispecified, which might explain poor estimation.'}
alpha0true <- as.data.frame(simulated_ltjmm$alpha0) %>% mutate(id = 1:n, parameter='alpha0') %>% 
  gather(outcome, truth, V1:V4)
alpha1true <- as.data.frame(simulated_ltjmm$alpha1) %>% mutate(id = 1:n, parameter='alpha1') %>% 
  gather(outcome, truth, V1:V4)
alphapm <- data.frame(
  parameter.id.outcome = grep('alpha', row.names(fit.sum), value=TRUE), 
  posterior.mean = fit.sum[grepl('alpha', row.names(fit.sum)), 'mean']) %>% 
  separate(parameter.id.outcome, c('parameter', 'id', 'outcome', 'other')) %>% 
  mutate(id = as.numeric(id))

pd <- full_join(alpha0true, alpha1true, by=c('id', 'outcome', 'parameter', 'truth')) %>%
  mutate(outcome = gsub('V', '', outcome)) %>% 
  full_join(alphapm) 
ggplot(pd, aes(x=truth, y=posterior.mean)) + 
  geom_point(aes(shape=parameter, color=parameter), alpha=0.25) +
  geom_abline(intercept=0, slope=1) + facet_wrap(~outcome, scales='free')
```

# Fit mixed effect model (MM) with independent univariate random effects with Stan
```{r stanFitMM, eval=FALSE}
fit <- ltjmm_stan(Y ~ year |
    1  | # fixed effects direct on outcome
    id | outcome,
  lt = FALSE,
  random_effects = "univariate",
  data = dd,
  seed = rng_seed,
  chains = 2, parallel_chains = 2,
  iter_warmup = 250, iter_sampling = 250, thin = 1)
# fit$save_object(file = 'sim_mm_results.RDS')
fit.sum <- fit$summary(c("beta", "gamma", "alpha0", "alpha1")) %>%
  as.data.frame()
row.names(fit.sum) <- fit.sum %>% pull(variable)
draws.beta <- fit$draws("beta")
draws.gamma <- fit$draws("gamma")
save(fit.sum, draws.beta, draws.gamma, 
  file = 'sim_mm_results.rdata')
```

```{r loadResultsMM}
load(system.file("testdata", "sim_mm_results.rdata", package="ltjmm"))
```

## Diagnostic plots
```{r traceplotBetaMM, fig.cap='Note. Model is mispecified, which might explain poor mixing.'}
mcmc_trace(draws.beta)
```

```{r traceplotGammaMM, fig.cap='Note. Model is mispecified, which might explain poor mixing.'}
mcmc_trace(draws.gamma)
```

## True versus posterior mean of random intercepts and slopes for each outcome
```{r biasAlphaPlotMM, fig.cap='Note. Model is mispecified, which might explain poor estimation.'}
alpha0true <- as.data.frame(simulated_ltjmm$alpha0) %>% mutate(id = 1:n, parameter='alpha0') %>% 
  gather(outcome, truth, V1:V4)
alpha1true <- as.data.frame(simulated_ltjmm$alpha1) %>% mutate(id = 1:n, parameter='alpha1') %>% 
  gather(outcome, truth, V1:V4)
alphapm <- data.frame(
  parameter.id.outcome = grep('alpha', row.names(fit.sum), value=TRUE), 
  posterior.mean = fit.sum[grepl('alpha', row.names(fit.sum)), 'mean']) %>% 
  separate(parameter.id.outcome, c('parameter', 'id', 'outcome', 'other')) %>% 
  mutate(id = as.numeric(id))

pd <- full_join(alpha0true, alpha1true, by=c('id', 'outcome', 'parameter', 'truth')) %>%
  mutate(outcome = gsub('V', '', outcome)) %>% 
  full_join(alphapm) 
ggplot(pd, aes(x=truth, y=posterior.mean)) + 
  geom_point(aes(shape=parameter, color=parameter), alpha=0.25) +
  geom_abline(intercept=0, slope=1) + facet_wrap(~outcome, scales='free')
```
